|

### 2024-10-21 manuscript/paper.Rmd
Update Manuscript
YS. editing, revising, and adding to the manuscript.

THIS FIGURE DOESN'T SHOW MUCH. LET'S CUT FOR SPACE. 
```{r fig.cap = "Fossil sample size for each time period"}
ggplot(aes(x = time, y = count), data = fortab) + geom_col() + xlab("Sample")  + scale_x_continuous(breaks = 1:18, labels = 1:18)
```

| Trait Name           | Trait Code | Trait Description           |
|----------------------|------------|-----------------------------|
|Standard Length|	`stl` |	Distance from anterior tip of premaxilla to posterior end of last vertebra (hypural plate)|
|Dorsal Spine | `mds` |	Number of dorsal spines from 0 to 3|
|Dorsal Fin Ray|`mdf`|	Number of bones in the dorsal fin posterior to the third dorsal spine (i.e., soft dorsal fin rays)
|Anal Fin Ray |`maf`|	Number of bones in the anal fin posterior to the anal spine (i.e., soft dorsal fin rays)|
|Abdominal Vertebra|mav|	Number of vertebrae anterior to the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Caudal Vertebra|`mcv`	| Number of vertebrae posterior to and including the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Pterygiophore number|`mpt`|	Number of pterygiophores anterior to but excluding the pterygiophore under the third dorsal spine, which is immediately anterior to and contiguous with the dorsal fin|
|Pelvic Spine length|`lps.sc`	|Length from the base of one pelvic spine above its articulation with the pelvic girdle to its distal tip|
|Ectocoracoid|`ect.sc`|	Length between the anterior and posterior tips of the shoulder girdle base (i.e., ectocoracoid)|
|Pelvic Girdle|`tpg.sc`|	Length between the anterior to posterior tips along midline. If vestigial, the sum of longest anterior-posterior axis for the vestiges|
|Cleithrum length|`cle.sc`|	Length from free dorsal tip to ventral tip of the cleithrum on the anterior margin of the shoulder girdle (i.e., cleithrum)|
|Premaxilla|`pmx.sc`|	Length from the anterior tip of the premaxilla to the distal tip of the ascending process of the premaxilla |
|Dorsal Spine|`Ds#.sc# = 1,2,or 3`|	Length from the base of a dorsal spine above the pterygiophore to its distal tip along the anterior edge|
|Pterygiophore|`lpt.sc`|	Distance between the anterior to posterior tips of the pterygiophore immediately preceding the 3rd dorsal spine (when present)|

Table: Traits and trait descriptions. ‘sc’ denotes size correction of trait against standard length. Names of bones follow @Bowne1994 unless otherwise noted.

<!-- I removed the table and replaced it with a figure -->
<!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
<!-- % Mon May  1 11:20:28 2023 -->
<!-- \begin{table}[ht] -->
<!-- \centering -->
<!-- \caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.} -->
<!-- \begin{tabular}{rl} -->
<!--   \hline -->
<!-- time & count \\  -->
<!--   \hline -->
<!--   1 & 43 \\  -->
<!--     2 & 41 \\  -->
<!--     3 & 51 \\  -->
<!--     4 & 41 \\  -->
<!--     5 & 46 \\  -->
<!--     6 & 48 \\  -->
<!--     7 & 67 \\  -->
<!--     8 & 55 \\  -->
<!--     9 & 42 \\  -->
<!--    10 & 33 \\  -->
<!--    11 & 37 \\  -->
<!--    12 & 22 \\  -->
<!--    13 & 41 \\  -->
<!--    14 & 43 \\  -->
<!--    15 & 46 \\  -->
<!--    16 & 47 \\  -->
<!--    17 & 56 \\  -->
<!--    18 & 55 \\  -->
<!--    \hline -->
<!-- \end{tabular} -->
<!-- \label{tabn} -->

## Extant Specimen data
To span the gamut of stickleback diversity for our predictive imputation model, we sampled modern stickleback from lakes containing generalist stickleback populations (@Hendryetal2009; @Bolnick2011) and from lakes containing benthic-limnetic species pairs (@Baumgartneretal1988; @SchluterandMcPhail1992). The generalist specimens used here were collected by YES in 2013 and were previously described in @Stuartetal2017 (Table S1). These samples were fixed in formalin, then stained for bone with Alizarin Red in 2013. Benthic and limnetic specimens were kindly loaned by D. Schluter and S. Blain at the University of British Columbia. The Schluter lab collected benthic and limnetic individuals from Enos Lake in 1988 and from Emily Lake, Little Quarry Lake, Paxton Lake, and Priest Lake in 2018 (Table S1). The Enos specimens had been fixed whole in formalin and stored in 40% isopropanol. The specimens from the other lakes were initially preserved whole in 95% ethanol in the field before being gradually transferred to water then formalin in the lab and ultimately stored in 40% isopropanol. In 2019, we stained these specimens for bone using Alizarin Red.
We next replicated fossil data collection (Table 1) on these extant specimens. Standard length as well as pelvic-spine length on each side were measured with calipers. We used a dissection microscope to count dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays. Right and left-side pelvic girdle lengths and ectocoracoid lengths were measured from ventral photographs taken using a Canon EOS Rebel T7 with a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy stand. Specimens were held in place for ventral photographs using a small tabletop vise with an attached scale bar. Lateral X-rays were used to measure dorsal spine length, number of pterygiophores anterior to the pterygiophore holding the third spine, length of the pterygiophore just anterior to the third spine, cleithrum length, and pre-maxilla ascending branch length. We also counted vertebrae from the X-rays: abdominal vertebrae were counted anterior to the first vertebra with a haemal spine contacting an anal fin pterygiophore. Caudal vertebrae were posterior, including the first vertebra with the haemal spine contacting the anal fin pterygiophore (following @Aguirreetal2014). X-rays were taken with an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the Field Museum of Natural History. Specimens were exposed at 35kV and 4mA. Small fish were exposed for 7s, medium fish for 8s, and large fish for 10s. We developed the film and scanned individual images of each fish using the B&W Negatives setting on an Epson Perfection 4990 Photo flatbed at 2400 dpi. Measurements from photographs and X-rays were taken with FIJI (@Schindelinetal2012) and its plugin ObjectJ (https://sils.fnwi.uva.nl/bcb/objectj/).

We dissected individuals from the generalist populations (Table S1) to determine sex from the gonads. Individuals from the species-pair lakes (TableS1) were previously sexed by the Schluter lab, using either dissection or a genotyping protocol (Whom, personal communication).

The extant data used here consists of a total of `r sum(!is.na(stick$gender))` specimens all with known sex (Table S1).  Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.

## Outlier analysis and size correction.

To check for outliers, we calculated within-group means and standard deviations for each trait separately for K series fossil specimens (pooled across samples) and for extant specimens (within generalist, benthic, or limnetic categories). We noted trait values greater than 3.0 standard deviations from the mean as potential outliers. We deemed 3.0 s.d. to be a reasonable threshold for detecting errors without excluding biologically relevant values. We checked whether these potential outliers were a result of data entry and collection error and corrected them if they were. We turned the remaining outlier trait values to NAs. We size-corrected continuous traits only, as they varied with size, unlike count traits that are fixed during early development. We regressed each trait on standard length using a mixed-model regression, pooling all specimens, following @Stuartetal2017. We appended the size corrected data to the uncorrected trait data frame and used all of our data (raw and size-corrected) to build the imputation model described next.

## Missing data imputation, including fossil sex
Briefly, We used multiple imputation (@little2002statistical) to impute the sex of the fossils. Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate sex of the stickleback fish and $\boldsymbol{Y}$ be an $(n_{extant} + n_{fossil}) \times K$ matrix of the $K$ phenotypes of interest.  Because the sex of the fossilized stickleback fish is unobservable, we further define $\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T$ where $\boldsymbol{W}_{extant}$ and $\boldsymbol{W}_{fossil}$ are the $n_{extant} \times 1$ and $n_{fossil} \times 1$ vectors of the observed extant sex and missing fossil sex, respectively. We imputeD missing sex for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching, implemented in R (CITATION).  Traditionally, the choice for the number of completed dats sets is a relatively small number such as $M = 5$ or $M = 10$.  However, @ZhouReiter2010 recommend a larger number of imputed data sets if users intend on performing Bayesian analysis after imputation.  Therefore, we imputed $M = 100$ complete datasets.

\textcolor{red}{presumably this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
\textcolor{red}{AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
\textcolor{red}{THE POOLING DESCRIBED IN THE NEXT SENTENCE. WAS IT DONE, AND IF SO, HOW WAS IT USED IN DOWNSTREAM ANALYSES. MATT SAYS THAT HE RAN IS MODELS ON EACH IMPUTED DATA SET. -YS}
We then pooled all of the draws from the posterior distribution across all the imputed data sets to estimate the posterior distributions of our parameters of interest (i.e., the trait values), rather than using Rubin's combining rules, following @ZhouReiter2010. We used these posterior distributions in our Bayesian analysis.

## Modeling trait means through time for males and females
The goal of our analysis was to estimate fossil trait means through time, by sex, while incorporating uncertainty in the imputed data, especially uncertainty sex classification. We took a Bayesian approach because imputation is the process of sampling from a posterior predictive distribution of missing data, given observed data. \textcolor{red}{THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH. -YS}

\textcolor{red}{DOES THE NEXT SENTENCE CONTRADICT THE PREVIOUS ONE? BELOW, WE SAY THAT WE USED EACH IMPUTED DATA SET. ABOVE, WE SAY THAT WE POOLED ACROSS IMPUTED DATA SETS.-YS}
To estimate fossil sample means through time, we first dropped the extant individuals from our imputed data sets. Then, for a given imputed dataset, we let $W_{ij}$ be the imputed sex and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fossil $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$.  In addition, we denote $Y_{K,ij}$, the last variable in $\boldsymbol{Y}_{ij}$, to be the standard length of each fish individual, a measure of body size.
\begin{align}
{Y}_{K,ij} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:stl}
\end{align}

It is reasonable to assume that the other continuous traits will have some correlation with standard length \textbf{(CITATION)}. We account for this by adding an additional parameter, $\gamma_k$, If $Y_{k,ij}$ is a continuous trait, then
\begin{align}
{Y}_{k,ij} & \iid \left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:cont}
\end{align}

If $Y_{k,ij}$ is a discrete trait, the conventional modelling method is to fit a Poisson distribution. However, because the empirical fossil data violate the assumption that the Poisson variance is equal to its mean, we fit discrete traits to a generalized Poisson model as defined in @GeneralizedPoisson. Specifically, if $X \sim GP(\lambda,\alpha)$, then 
\begin{align}
P(X = x) = \left\{\begin{array}{cc} \frac{(1 - \alpha)\lambda[(1 - \alpha)\lambda + \alpha x]^{x - 1} \exp\left\{-((1 - \alpha)\lambda  + \alpha x)\right\}}{x!} & (1 - \alpha)\lambda  + \alpha x \geq 0  \\ 0 & (1 - \alpha)\lambda  + \alpha x < 0 \end{array}\right.,
\label{eq:GP_pmf}
\end{align}
where GP denotes a generalized Poisson distribution with $E(X) = \lambda$ and $Var(X) = \frac{\lambda}{(1 - \alpha)^2}$.

In addition, we allowed that two discrete traits, abdominal vertebrae number (mav) and caudal vertebrae number (mcv) could also with  standard length (\textbf{CITATION}). This is because these are vertebral counts that run in parallel with the standard length measurement, along the anterior-posterior access of the fish; these vertebrae thus contribute to the length of the fish directly, and may evolve in a correlated fashion. If $Y_{k,ij}$ is one of the above traits, then
\begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_corr}
\end{align}

For the other discrete traits, we set $\gamma_k = 0$ because we expect that trait counts are set during early ontogeny, and should not change with size in adult fish. Thus,
\begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_ind}
\end{align}
We point out that, for the discrete traits, the means are represented by $\exp\{\mu_{k,ft_i}\}$ and $\exp\{\mu_{k,mt_i}\}$ for ease of modeling.

In the above model descriptions, $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$ model the time-$t_i$, sample-specific trait mean $k$ for females and males, respectively.  To account for the possibility that there is a time -dependent trend in $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$, we further set
\begin{align}
\mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
\label{eq:mu}
\end{align}
for $g \in \{f,m\}$, where $\beta_{0,kg}$ and $\beta_{1,kg}$ are regression parameters of phenotype $Y_k$ for each sex, and $u_{k,gt_i}$ is the corresponding residual. To account for potential correlations between the residuals for a given trait $k$ and sex $g$, we fit an Ornstein-Uhlenback (OU) process (@OUProcess). More specifically, we defined $du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}$, the change in $u_k,gt$ for a given trait $k$ and sex $g$ over a miniscule time period $dt$. The OU process is defined as 
\begin{align}
du_{k,gt} = -\kappa_k u_{k,gt} dt + \tau_k dW_t,
\label{eq:OU_cont}
\end{align}
where $\kappa_k$ is a parameter associated with the correlation between $u_{k,gt}$ and $u_{k,g(t+dt)}$, $\tau_k$ is the standard deviation of the OU process, and $W_t$ is a standard Brownian motion. As shown in @OUProcess, the closed form solution for the stochastic differential equation in (\ref{eq:OU_cont}) is
\begin{align}
u_{k,gt_i} \iid \mathcal{N}\left(u_{k,gt_{i-1}}\exp\{-\kappa_k(t_{i} - t_{i-1})\} , \frac{\tau_k^2(1 - \exp\{-2\kappa_k(t_{i} - t_{i-1}))\}}{2\kappa_k}\right)
\label{eq:OU_sol}
\end{align}
for $i = 2,\cdots,T$. In a traditional OU process, the initial value $u_{k,gt_1}$ is assumed to be a (potentially unknown) constant. This unknown constant can therefore be considered a latent variable, further justifying our Bayesian approach, which better incorporates prior information and uncertainty from missing data imputation in the estimate of the variable.

@@

All models were built using @R2022language and (\textbf STAN)

@@

# Results {#sec:results}
\textcolor{red}{WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS}

@@

>>>>

| Trait Name  | Trait Code | Trait Description           |
|---------|-----|------------------------------------------|
|Standard Length|	`stl` |	Distance from anterior tip of premaxilla to posterior end of last vertebra (hypural plate)|
|Dorsal Spine | `mds` |	Number of dorsal spines from 0 to 3|
|Dorsal Fin Ray|`mdf`|	Number of bones in the dorsal fin posterior to the third dorsal spine (i.e., soft dorsal fin rays)
|Anal Fin Ray |`maf`|	Number of bones in the anal fin posterior to the anal spine (i.e., soft dorsal fin rays)|
|Abdominal Vertebra|mav|	Number of vertebrae anterior to the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Caudal Vertebra|`mcv`	| Number of vertebrae posterior to and including the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Pterygiophore number|`mpt`|	Number of pterygiophores anterior to but excluding the pterygiophore under the third dorsal spine, which is immediately anterior to and contiguous with the dorsal fin|
|Pelvic Spine length|`lps.sc`	|Length from the base of one pelvic spine above its articulation with the pelvic girdle to its distal tip|
|Ectocoracoid|`ect.sc`|	Length between the anterior and posterior tips of the shoulder girdle base (i.e., ectocoracoid)|
|Pelvic Girdle|`tpg.sc`|	Length between the anterior to posterior tips along midline. If vestigial, the sum of longest anterior-posterior axis for the vestiges|
|Cleithrum length|`cle.sc`|	Length from free dorsal tip to ventral tip of the cleithrum on the anterior margin of the shoulder girdle (i.e., cleithrum)|
|Premaxilla|`pmx.sc`|	Length from the anterior tip of the premaxilla to the distal tip of the ascending process of the premaxilla |
|Dorsal Spine|`Ds#.sc #=1/2/3`|	Length from the base of a dorsal spine above the pterygiophore to its distal tip along the anterior edge|
|Pterygiophore|`lpt.sc`|	Distance between the anterior to posterior tips of the pterygiophore immediately preceding the 3rd dorsal spine (when present)|

\textcolor{red}{This table caption should be part of the table. Top row. at the very least, it should be the width of the table. -YS}
Table: Traits and trait descriptions. ‘sc’ denotes size correction of trait against standard length. Names of bones follow @Bowne1994 unless otherwise noted.

## Extant Specimen data
To span the gamut of stickleback diversity for our predictive imputation model, we sampled modern stickleback from lakes containing generalist stickleback populations (@Hendryetal2009; @Bolnick2011) and from lakes containing benthic-limnetic species pairs (@Baumgartneretal1988; @SchluterandMcPhail1992). The generalist specimens used here were collected by YES in 2013 and were previously described in @Stuartetal2017 (Table S1). These samples were fixed in formalin, then stained for bone with Alizarin Red in 2013. Benthic and limnetic specimens were kindly loaned by D. Schluter and S. Blain at the University of British Columbia. The Schluter lab collected benthic and limnetic individuals from Enos Lake in 1988 and from Emily Lake, Little Quarry Lake, Paxton Lake, and Priest Lake in 2018 (Table S1). The Enos specimens had been fixed whole in formalin and stored in 40% isopropanol. The specimens from the other lakes were initially preserved whole in 95% ethanol in the field before being gradually transferred to water then formalin in the lab and ultimately stored in 40% isopropanol. In 2019, we stained these specimens for bone using Alizarin Red.
We next replicated fossil data collection (Table 1) on these extant specimens. Standard length as well as pelvic-spine length on each side were measured with calipers. We used a dissection microscope to count dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays. Right and left-side pelvic girdle lengths and ectocoracoid lengths were measured from ventral photographs taken using a Canon EOS Rebel T7 with a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy stand. Specimens were held in place for ventral photographs using a small tabletop vise with an attached scale bar. Lateral X-rays were used to measure dorsal spine length, number of pterygiophores anterior to the pterygiophore holding the third spine, length of the pterygiophore just anterior to the third spine, cleithrum length, and pre-maxilla ascending branch length. We also counted vertebrae from the X-rays: abdominal vertebrae were counted anterior to the first vertebra with a haemal spine contacting an anal fin pterygiophore. Caudal vertebrae were posterior, including the first vertebra with the haemal spine contacting the anal fin pterygiophore (following @Aguirreetal2014). X-rays were taken with an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the Field Museum of Natural History. Specimens were exposed at 35kV and 4mA. Small fish were exposed for 7s, medium fish for 8s, and large fish for 10s. We developed the film and scanned individual images of each fish using the B&W Negatives setting on an Epson Perfection 4990 Photo flatbed at 2400 dpi. Measurements from photographs and X-rays were taken with FIJI (@Schindelinetal2012) and its plugin ObjectJ (https://sils.fnwi.uva.nl/bcb/objectj/). We dissected individuals from the generalist populations (Table S1) to determine sex from the gonads. Individuals from the species-pair lakes (TableS1) were previously sexed by the Schluter lab, using either dissection or a genotyping protocol \textcolor{red}{(Whom, personal communication)}.

## Outlier analysis and size correction. 
To check for outliers, we calculated within-group means and standard deviations for each trait separately for K series fossil specimens (pooled across samples) and for extant specimens (within generalist, benthic, or limnetic categories). We noted trait values greater than 3.0 standard deviations from the mean as potential outliers. We deemed 3.0 s.d. to be a reasonable threshold for detecting errors without excluding biologically relevant values. We checked whether these potential outliers were a result of data entry and collection error and corrected them if they were. We turned the remaining outlier trait values to NAs. We size-corrected continuous traits only, as they varied with size, unlike count traits that are fixed during early development. We regressed each trait on standard length using a mixed-model regression, pooling all specimens, following @Stuartetal2017. We appended the size corrected data to the uncorrected trait data frame and used all of our data (raw and size-corrected) to build the imputation model described next.

## Missing data imputation, including fossil sex
Briefly, we used multiple imputation (@little2002statistical) to impute the sex of the fossils. Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate sex of the stickleback fish and $\boldsymbol{Y}$ be an $(n_{extant} + n_{fossil}) \times K$ matrix of the $K$ phenotypes of interest.  Because the sex of the fossilized stickleback fish is unobservable, we further define $\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T$ where $\boldsymbol{W}_{extant}$ and $\boldsymbol{W}_{fossil}$ are the $n_{extant} \times 1$ and $n_{fossil} \times 1$ vectors of the observed extant sex and missing fossil sex, respectively. We imputed missing sex for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching, implemented in R (CITATION).  Traditionally, the choice for the number of completed dats sets is a relatively small number such as $M = 5$ or $M = 10$.  However, @ZhouReiter2010 recommend a larger number of imputed data sets if users intend on performing Bayesian analysis after imputation.  Therefore, we imputed $M = 100$ complete datasets. \textcolor{red}{PRESUMABLY this also imputes missing data in the other traits? We should probably say that explicitly. -YS}

\textcolor{red}{AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}

We then pooled all of the draws from the posterior distribution across all the imputed data sets to estimate the posterior distributions of our parameters of interest (i.e., the trait values), rather than using Rubin's combining rules, following @ZhouReiter2010. We then dropped extant fish from our imputed data sets and used these posterior distributions in our downstream analyses of fossil sexual dimorphism.

## Modeling fossil stickleback trait means through time for males and females
The goal of our analysis was to estimate fossil trait means through time, by sex, while incorporating uncertainty across imputations in the classification of sex. We used a Bayesian approach to estimate trait means in part because imputation is the process of sampling from a posterior predictive distribution of missing data, given observed data. \textcolor{red}{THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH, EXPLAINING IT TO MY AUDIENCE. -YS}

To estimate fossil sample means through time, for a given imputed dataset, we let $W_{ij}$ be the imputed sex and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fossil $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$.  In addition, we denote $Y_{K,ij}$, the last variable in $\boldsymbol{Y}_{ij}$, to be the standard length of each fish individual, a measure of body size.
\begin{align}
{Y}_{K,ij} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:stl}
\end{align}

Continuous traits will likely have some correlation with standard length (i.e., allometry) (@Huxley1932)(@Vojeetal2022)(@Vojeetal2014). We account for this by adding an additional parameter, $\gamma_k$, If $Y_{k,ij}$ is a continuous trait, then
\begin{align}
{Y}_{k,ij} & \iid \left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:cont}
\end{align}

If $Y_{k,ij}$ is a discrete trait, the conventional modelling method is to fit a Poisson distribution. However, the empirical fossil data violate the assumption that the Poisson variance is equal to its mean, so we fitted discrete traits to a generalized Poisson model (@GeneralizedPoisson). Specifically, if $X \sim GP(\lambda,\alpha)$, then 
\begin{align}
P(X = x) = \left\{\begin{array}{cc} \frac{(1 - \alpha)\lambda[(1 - \alpha)\lambda + \alpha x]^{x - 1} \exp\left\{-((1 - \alpha)\lambda  + \alpha x)\right\}}{x!} & (1 - \alpha)\lambda  + \alpha x \geq 0  \\ 0 & (1 - \alpha)\lambda  + \alpha x < 0 \end{array}\right.,
\label{eq:GP_pmf}
\end{align}
where GP denotes a generalized Poisson distribution with $E(X) = \lambda$ and $Var(X) = \frac{\lambda}{(1 - \alpha)^2}$.

In addition, we modeled two discrete traits, abdominal vertebrae number (mav) and caudal vertebrae number (mcv) to vary with standard length. This is because these are vertebral counts that made along the anterior-posterior access of the fish; these vertebrae contribute to the length of the fish directly and may evolve in a correlated fashion. If $Y_{k,ij}$ is one of the above traits, then
\begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_corr}
\end{align}

For the other discrete counting traits, we set $\gamma_k = 0$ because we expect that trait counts are set during early ontogeny and should not change with size in adult fish. Thus,
\begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_ind}
\end{align}
We point out that counting trait means are represented by $\exp\{\mu_{k,ft_i}\}$ and $\exp\{\mu_{k,mt_i}\}$ for ease of modeling.

In the above model descriptions, $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$ model the time-$t_i$, sample-specific trait mean $k$ for females and males, respectively.  To account for the possibility that there is a time-dependent (i.e., evolutionary trend) trend in $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$, we further set
\begin{align}
\mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
\label{eq:mu}
\end{align}
for $g \in \{f,m\}$, where $\beta_{0,kg}$ and $\beta_{1,kg}$ are regression parameters of phenotype $Y_k$ for each sex, and $u_{k,gt_i}$ is the corresponding residual. To account for potential correlations between the residuals for a given trait $k$ and sex $g$, we fit an Ornstein-Uhlenback (OU) process (@OUProcess). More specifically, we defined $du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}$, the change in $u_k,gt$ for a given trait $k$ and sex $g$ over a miniscule time period $dt$. The OU process is defined as 
\begin{align}
du_{k,gt} = -\kappa_k u_{k,gt} dt + \tau_k dW_t,
\label{eq:OU_cont}
\end{align}
where $\kappa_k$ is a parameter associated with the correlation between $u_{k,gt}$ and $u_{k,g(t+dt)}$, $\tau_k$ is the standard deviation of the OU process, and $W_t$ is a standard Brownian motion. As shown in @OUProcess, the closed form solution for the stochastic differential equation in (\ref{eq:OU_cont}) is
\begin{align}
u_{k,gt_i} \iid \mathcal{N}\left(u_{k,gt_{i-1}}\exp\{-\kappa_k(t_{i} - t_{i-1})\} , \frac{\tau_k^2(1 - \exp\{-2\kappa_k(t_{i} - t_{i-1}))\}}{2\kappa_k}\right)
\label{eq:OU_sol}
\end{align}
for $i = 2,\cdots,T$. In a traditional OU process, the initial value $u_{k,gt_1}$ is assumed to be a (potentially unknown) constant. This unknown constant can therefore be considered a latent variable, further justifying our Bayesian approach, which better incorporates prior information and uncertainty from missing data imputation in the estimate of latent variables.

@@

@@

# Results {#sec:results}
\textcolor{red}{WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS}
The extant data used here consists of a total of `r sum(!is.na(stick$gender))` specimens all with known sex (Table S1).  Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.

@@

<!-- SCRAP CODE and TEXT. YS moved it down here on 21 October 2024 because it was making it hard for me to see the manuscript.  -->
<!-- THIS FIGURE DOESN'T SHOW MUCH. LET'S CUT FOR SPACE.  -->
```{r fig.cap = "Fossil sample size for each time period",include=FALSE}
ggplot(aes(x = time, y = count), data = fortab) + geom_col() + xlab("Sample")  + scale_x_continuous(breaks = 1:18, labels = 1:18)
```
<!-- I removed the table and replaced it with a figure -->
<!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
<!-- % Mon May  1 11:20:28 2023 -->
<!-- \begin{table}[ht] -->
<!-- \centering -->
<!-- \caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.} -->
<!-- \begin{tabular}{rl} -->
<!--   \hline -->
<!-- time & count \\  -->
<!--   \hline -->
<!--   1 & 43 \\  -->
<!--     2 & 41 \\  -->
<!--     3 & 51 \\  -->
<!--     4 & 41 \\  -->
<!--     5 & 46 \\  -->
<!--     6 & 48 \\  -->
<!--     7 & 67 \\  -->
<!--     8 & 55 \\  -->
<!--     9 & 42 \\  -->
<!--    10 & 33 \\  -->
<!--    11 & 37 \\  -->
<!--    12 & 22 \\  -->
<!--    13 & 41 \\  -->
<!--    14 & 43 \\  -->
<!--    15 & 46 \\  -->
<!--    16 & 47 \\  -->
<!--    17 & 56 \\  -->
<!--    18 & 55 \\  -->
<!--    \hline -->
<!-- \end{tabular} -->
<!-- \label{tabn} -->
